A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-03 podc_paper/correctness.tex
working on proof

\begin{invariant}\label{lem::headPosition} 
For $0 \leq i < v.\head$, $v.\fld{blocks}[i]\neq\nl$.  For $i>v.\head$, $v.\fld{blocks}[i]=\nl$.
\end{invariant}

\begin{proof}
Initially, $v.\head=1$, $v.\fld{blocks}[0]\neq\nl$  and $v.\fld{blocks}[i]=\nl$ for  $i>0$, so the claim~holds.

@@

Now, assume the claim holds before a change to $v.\head$, which can only be an increment from $h$ to $h+1$
by a successful \op{CAS} at line \ref{incrementHead} of \op{Advance}.
It suffices to show that $v.\fld{blocks}[head] \neq \nl$.
\nf{Advance} is called either at line \ref{helpAdvance} 
after testing that $v.\fld{blocks}[h]\neq\nl$ at line \ref{ifHeadnotNull},
or at line \ref{advance} after the \op{CAS} at line \ref{cas} ensures $v.\fld{blocks}[h]\neq\nl$. 
\end{proof}

@@

-------

\subsection{Correctness of IndexDequeue}
The next few results show that the \nf{super} field of a block is accurate within one of the actual index of the block's superblock in the parent node. 
\here{mention that precondition of IndexDequeue follows; also check that block of parent indexed by sup in that routine is non-null}
Then we explain how it is used to compute the rank of a given \nf{Dequeue} in the root.
\begin{definition}\label{orderRefresh} If a \nf{Refresh} instance $R_1$ does its \nf{CAS} at Line \ref{cas} earlier than \nf{Refresh} instance $R_2$ we say $R_1$ has \it{happened before} $R_2$.
\end{definition}

\begin{observation} \label{setSupBeforeIncHead}
After \nf{$n$.blocks[$i$].CAS(null, $B$)} succeeds, \nf{$n$.head} cannot increase from ${i}$ to ${i+1}$ until \nf{$B$.super} is set.
\end{observation}
\begin{proof}
From Observation \ref{nonDecreasingHead} we know that \nf{$n$.head} changes only by the increment on Line \ref{incrementHead}. Before an instance of \nf{Advance} increments \nf{n.head} on Line \ref{incrementHead}, Line \ref{setSuper1} ensures that \nf{n.blocks[head].super} was set at Line \ref{setSuper1}.
\end{proof}

\begin{corollary}
  If \nf{$n$.blocks[$i$].super} is \nf{null}, then $\nf{$n$.head}\leq i$ and \nf{$n$.blocks[$i+1$]} is \nf{null}.
\end{corollary}
\begin{proof}
By Invariant \ref{lem::headPosition} and Observation \ref{setSupBeforeIncHead}.
\end{proof}
Now let us consider how the \nf{Refresh}es that took place on the parent of node $n$ after block $B$ was stored in $n$ will help to set \nf{$B$.super} and propagate $B$ to the parent.

@@

\begin{lemma}
  If $B=n$\nf{.blocks[}$b$\nf{]} is a direct subblock of \nf{$n$.parent.blocks[$superblock$]} then $\nf{$B$.super}\leq superblock$.
\end{lemma}
\begin{proof}
Let $R_p$ be the instance of $n$\nf{.parent.Refresh} that does a successful \nf{CAS}(Line~\ref{cas}) and puts the superblock of $B$ which is $n$\nf{.parent.blocks[$superblock$]} into $n\nf{.parent}$. By Observation~\ref{lem::createBlockHead} if $R_p$ propagates $B$ it has to read a greater value than $b$ from \nf{$n$.head}, which means \nf{$n$.head} was incremented from $b$ to $b+1$ in Line \ref{incrementHead}. By Observation \ref{setSupBeforeIncHead} \nf{$B$.super} was already set in Line \ref{setSuper1}. The value written in \nf{$B$.super}, was read in Line \ref{readParentHead} before the \nf{CAS} that sets \nf{$B$.super} in Line \ref{setSuper1}. From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $B\nf{.super}\leq superblock$, since \nf{$n$.parent.head} is still equal to $superblock$ when $R_p$ executes its \nf{CAS} at Line \ref{cas} by Lemma \ref{lem::headProgress}. The reader may wonder when the case $b\nf{.super}=superblock$ happens. This can happen when $\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.
\end{proof}

@@

We prove \nf{IndexDequeue}'s correctness using Corollary \ref{superRelation} on each step of the \nf{IndexDequeue}.

>>>>

\begin{invariant}\label{lem::headPosition} 
For $0 \leq i < v.\head$, $v.\fld{blocks}[i]\neq\nl$.  For $i>v.\head$, $v.\fld{blocks}[i]=\nl$.
If $v\neq \var{root}$,  $v.\fld{blocks}[i].super \neq \nl$ for $0<i<v.head$.
\end{invariant}

\begin{proof}
Initially, $v.\head=1$, $v.\fld{blocks}[0]\neq\nl$  and $v.\fld{blocks}[i]=\nl$ for  $i>0$, so the claims~hold.

@@

Now, assume the claim holds before a change to $v.\head$, which can only be an increment from $h$ to $h+1$
by a successful \op{CAS} at line \ref{incrementHead} of \op{Advance}.
For the first two claims, it suffices to show that $v.\fld{blocks}[head] \neq \nl$.
\nf{Advance} is called either at line \ref{helpAdvance} 
after testing that $v.\fld{blocks}[h]\neq\nl$ at line \ref{ifHeadnotNull},
or at line \ref{advance} after the \op{CAS} at line \ref{cas} ensures $v.\fld{blocks}[h]\neq\nl$.
For the third claim, observe that prior to incrementing $v.\head$ at line \ref{incrementHead},
the \op{CAS} at line \ref{setSuper1} ensures that $v.\fld{blocks}[i].super\neq \nl$.
\end{proof}

@@

\subsection{Correctness of IndexDequeue}

The next few results show that the \nf{super} field of a block is accurate within one of the actual index of the block's superblock in the parent node.

\begin{lemma}
  If $B=v.\var{blocks}[b]$ is a direct subblock of $v.\fld{parent.blocks}[s]$ then $B.\fld{super}\leq s$.
\end{lemma}
-------

\begin{proof}
Let $R_p$ be the instance of $n$\nf{.parent.Refresh} that does a successful \nf{CAS}(Line~\ref{cas}) and puts the superblock of $B$ which is $n$\nf{.parent.blocks[$s$]} into $n\nf{.parent}$. By Observation~\ref{lem::createBlockHead} if $R_p$ propagates $B$ it has to read a greater value than $b$ from \nf{$n$.head}, which means \nf{$n$.head} was incremented from $b$ to $b+1$ in Line \ref{incrementHead}. By Observation \ref{setSupBeforeIncHead} \nf{$B$.super} was already set in Line \ref{setSuper1}. The value written in \nf{$B$.super}, was read in Line \ref{readParentHead} before the \nf{CAS} that sets \nf{$B$.super} in Line \ref{setSuper1}. From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $B\nf{.super}\leq s$, since \nf{$n$.parent.head} is still equal to $s$ when $R_p$ executes its \nf{CAS} at Line \ref{cas} by Lemma \ref{lem::headProgress}. The reader may wonder when the case $b\nf{.super}=s$ happens. This can happen when $\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.
\end{proof}

Then we explain how it is used to compute the rank of a given \nf{Dequeue} in the root.
\begin{definition}\label{orderRefresh} If a \nf{Refresh} instance $R_1$ does its \nf{CAS} at Line \ref{cas} earlier than \nf{Refresh} instance $R_2$ we say $R_1$ has \it{happened before} $R_2$.
\end{definition}

Now let us consider how the \nf{Refresh}es that took place on the parent of node $n$ after block $B$ was stored in $n$ will help to set \nf{$B$.super} and propagate $B$ to the parent.

@@

@@

We prove \nf{IndexDequeue}'s correctness using Corollary \ref{superRelation} on each step of the \nf{IndexDequeue}.
\here{mention somewhere why precondition of IndexDequeue is true; also check that block of parent indexed by sup in that routine is non-null}
